---
title: "Gitsbe Cell Line Model Analysis"
output:
  html_document:
    df_print: paged
    #self_contained: no
---

## Intro

This notebook includes the ensemble model analysis performed on the models 
generated by the `Gitsbe` module when running the DrugLogics computational 
pipeline for finding synergistic drug combinations (drug pairs). All these 
models were trained towards a specific steady state signaling pattern that was
derived based on input data (gene expression, CNV) for a particular cell line. 
The input for the simulations and the output data are in the `cell-lines-2500` 
directory (the 2500 number denotes the number of simulations executed). The 
analysis will be presented step by step in the sections below.

The R version used for this analysis is: **3.4.4 (Someone to Lean On)**.

## Prerequisites

Firstly, we load the required libraries (you need to install them if you don't
have them):
```{r Load libraries, message=FALSE}
# require(RColorBrewer)
# require(heatmap.plus)
# require(gplots)
require(rje)
```
and the relevant helper functions:
```{r Helper functions}
# Set the working directory to the gitsbe-model-analysis folder: 
# setwd("pathTo/gitsbe-model-analysis")
source("Rscripts/input_functions.R")
source("Rscripts/output_functions.R")
source("Rscripts/analysis_functions.R")
source("Rscripts/plot_functions.R")
```

## Input

We will define the name of the cell line which must match the name of the 
directory that has the input files inside the `cell-lines-2500` directory. Our
analysis will be done on the `A498` cell line (though it is applicable with 
small technical changes to the other cell lines):
```{r Input: cell line}
cell.line = "A498"
data.dir = paste0(getwd(), "/cell-lines-2500/", cell.line, "/")

# create 'results' directory if it doesn't exist already 
results.dir = paste0(data.dir, "results")
dir.create(results.dir, showWarnings = FALSE)

print(paste("Cell line:", cell.line))
```
**Three inputs** are used in this analysis:

- The **model_predictions** file which has for each model the prediction for 
each drug combination tested (either *0* = no synergy predicted, *1* = synergy 
predicted, *NA* = couldn't find stable states in either the drug combination 
inhibited model or in any of the two single-drug inhibited models)
- The **observed_synergies** file which lists the drug combinations that were 
observed as synergistic for the particular cell line.
- The **models** directory, which is the same as the models directory produced 
by `Gitsbe` and has one `.gitsbe` file per model that includes this info:
    - The *fitness* to the steady state (value between 0 and 1)
    - The *stable state* of the boolean model. Note that a model can have 1 
stable state or none in our simulations - but the models used in this analysis 
have been selected through a genetic evolution algorithm in `Gitsbe` and so in 
the end, only those with 1 stable state have higher fitness values and remain in 
the later generations. Higher fitness here means a better match of a model's 
stable state to the cell line derived steady state (a perfect match would result 
in a fitness of 1)
    - The *boolean equations* of the model

```{r Input: files}
model.predictions.file = paste0(data.dir, "model_predictions")
observed.synergies.file = paste0(data.dir, "observed_synergies")
models.dir = paste0(data.dir, "models")
```
Now, we parse the data into proper R objects. First the synergy predictions 
per model:
```{r Input: model predictions}
model.predictions = get.model.predictions(model.predictions.file)
head(model.predictions)
```
So, we can see that our dataset has the models as rows and each column is a 
different drug combination that was tested in our simulations.
```{r Model&Drug Stats}
drug.combinations.tested = colnames(model.predictions)
models                   = rownames(model.predictions)
nodes                    = get.node.names(models.dir)

number.of.drugs.tested = length(drug.combinations.tested)
number.of.models       = length(models)
number.of.nodes        = length(nodes)

print(paste("Drug combinations tested: ", number.of.drugs.tested, "",
            "Number of models: ", number.of.models, "",
            "Number of nodes:", number.of.nodes))
```
Next, we get the full stable state, the equations and fitness score per model:
```{r Input: models stable states}
models.stable.state = get.stable.state.from.models.dir(models.dir)
models.stable.state = models.stable.state[models,]
head(as.data.frame(models.stable.state))
```
The rows of the above dataset represent the models while the columns are the
names of the nodes (proteins, genes, etc.) of the cancer cell network under
study. So, each model has one stable state which means that in every model, the
nodes in the network have reached a state of either 0 (inhibition) or 1
(activation).
```{r Input: models equations}
models.equations = get.equations.from.models.dir(
  models.dir, remove.equations.with.link.operator = TRUE)
models.equations = models.equations[models,]
head(as.data.frame(models.equations))
```
For the equations, if we look at a specific row (a model so to speak), the 
columns (node names) correspond to the *targets of regulation* (and the network 
has been built so that *every node is a target* - i.e. it has other nodes 
activating or inhibiting it). The general form of a boolean equation is: 
**Target \*= (Activator OR Activator OR...) AND NOT (Inhibitor OR Inhibitor 
OR...)**.

The difference between the models' boolean equations is the *link operator* 
(OR NOT/AND NOT) which has been mutated (changed) through the evolutionary 
process of the genetic algorithm in `Gitsbe`. For example, if a model has for 
the column `ERK_f` a value of 1, the correspoding equation is: 
ERK_f \*= (MEK_f) **OR NOT** ((DUSP6) OR PPP1CA). A value of 0 would correspond 
to **AND NOT** in that case. Note that the equations that do not have link 
operators (meaning that they are *the same for every model*) are discarded 
(so less columns in this dataset) since we want to see the change/difference 
between the models.

We, also keep track of each model fitness to the steady state (a value closer 
to 1 corresponds to a better fit to the steady state profile):
```{r Input: models fitness}
models.fitness = get.fitness.from.models.dir(models.dir)
models.fitness = models.fitness[models]
head(as.data.frame(models.fitness))
```
Lastly, the synergies observed are:
```{r Input: observed synergies}
observed.synergies = get.observed.synergies(
  observed.synergies.file, drug.combinations.tested)
number.of.observed.synergies = length(observed.synergies)

print(paste("Number of synergies observed:", number.of.observed.synergies))
observed.synergies
```

## Analysis

It will be interesting to know the percentage of the above observed synergies 
that were actually predicted by at least one of the models (there might be 
combinations that no model out of the 7500 could predict):
```{r Synergy stats 1}
# Subset the model.data to the observed synergies
observed.model.predictions = 
  model.predictions[,sapply(drug.combinations.tested, function(drug.comb) {
    is.correct.synergy(drug.comb, observed.synergies)
  })]

# Subset the model.data to the unobserved synergies
unobserved.model.predictions = 
  model.predictions[,sapply(drug.combinations.tested, function(drug.comb) {
    !is.correct.synergy(drug.comb, observed.synergies)
  })]

stopifnot(dim(observed.model.predictions)[2] + 
          dim(unobserved.model.predictions)[2] == number.of.drugs.tested)

number.of.models.per.observed.synergy = 
  apply(observed.model.predictions, 2, sum, na.rm = T)
predicted.synergies = names(number.of.models.per.observed.synergy)[
                            number.of.models.per.observed.synergy > 0]
number.of.models.per.observed.synergy
predicted.synergies
print(paste0("Percentage of predicted synergies (by at least one model): ", 
      100 * length(predicted.synergies)/number.of.observed.synergies, "%"))
```
So, for the `A498` cell line, there were indeed synergies that no model could 
predict. Next, we want to know the maximum number of possible predicted 
synergies by one model alone - can one model by itself predict all the above 
predicted synergies? To do that, we go even further and **count the number of 
models that predict a specific set of synergies** for every possible combination 
subset of the predicted synergy set found above:
```{r Synergy stats 2}
# Find the number of predictive models for every synergy subset
predicted.synergies.powerset = powerSet(predicted.synergies)
predicted.synergies.powerset = predicted.synergies.powerset[
  order(sapply(predicted.synergies.powerset, length))
]
names(predicted.synergies.powerset) = 
  sapply(predicted.synergies.powerset, function(drug.comb.set) {
    paste(drug.comb.set, collapse = ",")
  })

synergy.subset.stats = 
  sapply(predicted.synergies.powerset, function(drug.comb.set) { 
    count.models.that.predict.synergy.set(drug.comb.set, observed.model.predictions) 
})

# output.vector.to.file(cell.line = "A498", "synergy_subsets", synergy.subset.stats, with.row.names = TRUE)
```



## Statistical Analysis - TODO
```{r}
# Count the predictions of the observed synergies per model (TP)
models.synergies.tp = apply(observed.model.predictions, 1, sum, na.rm = T)
model.synergies.stats = table(models.synergies.tp)

# Count the false negatives (FN)
models.synergies.fn = apply(observed.model.data, 1, function(x) {
  sum( x == 0 | is.na(x) )
})

# P = TP + FN (Positives)
positives = ncol(observed.model.data)
models.synergies.p = models.synergies.tp + models.synergies.fn
stopifnot(sum(models.synergies.p == positives) == length(models.synergies.p))

# Count the predictions of the non-observed synergies per model (FP)
models.synergies.fp = apply(unobserved.model.data, 1, sum, na.rm = T)

# Count the True Negatives (TN)
models.synergies.tn = apply(unobserved.model.data, 1, function(x) {
  sum( x == 0 | is.na(x))
})

# N = FP + TN (Negatives)
negatives = ncol(unobserved.model.data)
models.synergies.n = models.synergies.fp + models.synergies.tn
stopifnot(sum(models.synergies.n == negatives) == length(models.synergies.n))

# Calculate the Sensitivity/Recall (TPR), Specificity (TNR), 
# Precision (PPV) and False Positive Rate (FPR)
models.sensitivity = models.synergies.tp / models.synergies.p
models.specificity = models.synergies.tn / models.synergies.n
models.precision   = models.synergies.tp / 
  (models.synergies.tp + models.synergies.fp)
models.fpr         = models.synergies.fp / models.synergies.n

# save stats all together in one file
measure.stats = cbind(models.fitness, models.synergies.tp, 
                      models.synergies.fp, models.synergies.tn, 
                      models.synergies.fn, models.sensitivity, 
                      models.specificity, models.precision, models.fpr)
output.data.to.file(results.dir, 
                    paste("statistical_measures_", cell.line, ".txt", sep = ""),
                    measure.stats, with.col.names = TRUE)
```

```{r}
# ROC curve
roc = get.average.over.unique.values(models.fpr, models.sensitivity)
roc = rbind(c(0,0,0), roc, c(1,1,0))
plot(roc, xlim = c(0,1), ylim = c(0,1), type = "o", main = "ROC curve")
```

```{r}
# Next plot makes more sense:
fitness.vs.avg.fp = 
  get.average.over.unique.values(models.fitness, models.synergies.fp)
x = fitness.vs.avg.fp[,1]
avg = fitness.vs.avg.fp[,2]
sdev = fitness.vs.avg.fp[,3]
plot(fitness.vs.avg.fp, ylim=range(c(avg-sdev, avg+sdev)))
arrows(x, avg-sdev, x, avg+sdev, length=0.05, angle=90, code=3)
```

