---
title: "Data Preprocessing"
author: "[John Zobolas](https://github.com/bblodfon)"
output: 
  html_document:
    theme: united
    toc: true
    toc_float:
      collapsed: false
---

## Intro

The purpose of this R markdown document is to extract from one or multiple 
**models** directories the models' **stable states** and **boolean link 
operators** to proper R data structures and save them for later use. This data 
extraction can take some time and it is better to have the data in tab delimited
files, ready to be loaded as matrices before the analysis is done for each cell 
line.

## Data Structure and Output

The structure of the data and input to this document should be a 
single directory (`input.dir`). In this directory there could be one or many 
cell line directories, each one having a **models** directory (having all the 
models produced by the `Gitsbe` module in `.gitsbe` format), a **model_predictions** 
file, an **observed_synergies** file and maybe a **training_data** file. 

The output will be two more files in each cell line directory: **models_stable_state** 
and **models_equations**.

## Data extraction

```{r Load libraries, message = FALSE}
library(usefun)
library(emba)
```

```{r Input}
# give the relative path to the input data dir
input.dir = "/atopo/cell-lines-2500"
data.dir = paste0(getwd(), input.dir)

cell.line.dirs = list.dirs(path = data.dir, recursive = FALSE)
```

```{r Data extraction and saving}
for (cell.line.dir in cell.line.dirs) {
  cell.line = basename(cell.line.dir)
  print(paste0("Saving files for ", cell.line, " cell line..."))
  models.dir = paste0(cell.line.dir, "/models")
  
  model.predictions.file = paste0(cell.line.dir, "/model_predictions")
  models.stable.state.file = paste0(cell.line.dir, "/models_stable_state")
  models.equations.file = paste0(cell.line.dir, "/models_equations")
  
  # for having the row names in the same order
  model.predictions = get_model_predictions(model.predictions.file)
  models = rownames(model.predictions)
  
  models.stable.state = get_stable_state_from_models_dir(models.dir)
  models.stable.state = models.stable.state[models,]
  save_mat_to_file(mat = models.stable.state, file = models.stable.state.file)
  
  models.equations = get_link_operators_from_models_dir(models.dir)
  models.equations = models.equations[models,]
  save_mat_to_file(mat = models.equations, file = models.equations.file)
}
```
