---
title: "Gitsbe Model Prediction Performance Analysis"
output:
  html_document:
    df_print: paged
---

## Intro

This notebook includes the ensemble model analysis performed on the models generated by the Gitsbe module (part of the DrugLogics pipeline) when running the DrugLogics computational pipeline for finding synergistic drug combinations. The analysis will be presented step by step in the text below.

The R version used for this analysis is: **3.4.4 (Someone to Lean On)**.

## Prerequisites

Firstly, we load the required libraries (you need to install them if you don't have them):

```{r message=FALSE}
# require(RColorBrewer)
# require(heatmap.plus)
# require(gplots)
# require(rje)
```

and the relavant helper functions:

```{r}
setwd(".")
source("input_functions.R")
#source("output_functions.R")
#source("plot_functions.R")
#source("analysis_functions.R")
```

## Input

We will define the name of the cell line which must match the name of the directory that has the input files. There are **3 inputs** used in this analysis:

- The **model_predictions** file which has for each model the prediction for each synergy combination (either *0* = no synergy predicted, *1* = synergy predicted, *NA* = couldn't find stable states in either the drug combination or in any of the single drugs used in the model)
- The **observed_synergies** file which lists the drug combinations that were observed as synergistic.
- THe **models** directory, which is the same as the models directory produced by Gitsbe and has one file per model and includes this info:
    - The *fitness* to the steady state (value between 0 and 1)
    - The *stable state* of the boolean model (can be either 1 stable state or none - the later rarely happens if not at all, since the models have been selected through a genetic evolution algorithm in Gitsbe and so in the end, only those with at least one stable state have higher fitness values and stay in the later generations) and 
    - The *boolean equations* of the model

```{r}
cell.line = "AGS"
data.dir = paste(getwd(), "/", cell.line, "/", sep = "")
model.predictions.file = paste(data.dir, "model_predictions", sep = "")
observed.synergies.file = paste(data.dir, "observed_synergies", sep = "")
models.dir = paste(data.dir, "models", sep = "")
```

Now, we parse the data into proper R objects. First the synergy predictions per model:
```{r}
model.data = read.model.predictions.file(model.predictions.file)
head(model.data)
```
So, we can see that our dataset has the models as rows and each column is a different drug combination tested.
```{r}
drug.combinations.tested = colnames(model.data)
models                   = rownames(model.data)

models.number        = dim(model.data)[1]
drug.comb.number     = dim(model.data)[2]
print(paste("Drug combinations tested: ", drug.comb.number, ",",
            "Number of models: ", models.number))
```
 The synergies observed are:
```{r}
synergy.observations.data = read.observed.synergies.file(
  observed.synergies.file, drug.combinations.tested)
synergy.observations.data
```
Next, we get the full stable state, the equations and fitness value per model:
```{r}
models.stable.state = get.stableState.from.models.dir(models.dir)
models.stable.state = models.stable.state[models,]
head(as.data.frame(models.stable.state))
```
Here, the rows of these dataset represent the models while the columns are the names of the nodes (proteins, genes, etc.) of the cancer cell network under study. So, each model has one stable state which means that in every model, the nodes in the network have been assigned a value of either 0 (inhibition) or 1 (activation).
```{r}
models.equations = get.equations.from.models.dir(
  models.dir, removeEquationsHavingLinkOperator = TRUE)
models.equations = models.equations[models,]
head(as.data.frame(models.equations))
```
For the equations, if we look at a specific row (a model so to speak), the columns (node names) correspond to the *targets of regulation* (and the network has been built so that *every node is a target* - i.e. it has other nodes activating or inhibiting it). The general form of a boolean equation is: **Target \*= (Activator OR Activator OR...) AND NOT (Inhibitor OR Inhibitor OR...)**.

The difference between the models' boolean equations is the link operator (OR/AND) which has been mutated (changed) through the evolutionary process of the genetic algorithm in Gitsbe. For example, if a model has for the column ERK_f a value of 1, the correspoding equation is: ERK_f \*= (MEK_f) **OR NOT** ((DUSP6) OR PPP1CA). A value of 0 would correspond to **AND NOT** in that case. Note that the equations that do not have link operators (meaning that they are *the same for every model*) are discarded (so less columns in this dataset) since we want to see the change/difference between the models.

We, also keep track of each model fitness to steady state (a value close to 1 corresponds to a perfect fit to the steady state profile):
```{r}
models.fitness = get.fitness.from.models.dir(models.dir)
models.fitness = models.fitness[models]
head(as.data.frame(models.fitness))
```






