---
title: "MDA-MB-468 Cell Line Model Analysis"
author: "[John Zobolas](https://github.com/bblodfon)"
output:
  html_document:
    df_print: paged
    theme: united # also good: default, cosmo, readable
    toc: true
    toc_float: true
    toc_depth: 5
    number_sections: false
    #self_contained: no
---

## Intro

This notebook includes the ensemble model analysis performed on the models 
generated by the `Gitsbe` module when running the DrugLogics computational 
pipeline for finding synergistic drug combinations (drug pairs). All these 
models were trained towards a specific steady state signaling pattern that was
derived based on input data (gene expression, CNV) for the `MDA-MB-468` cell line, 
the use of the [PARADIGM](https://doi.org/10.1093/bioinformatics/btq182) software 
and a topology that was build for simulating a cancer cell fate decision network.
The input for the simulations and the output data are in the `cell-lines-2500` 
directory (the 2500 number denotes the number of simulations executed). The 
analysis will be presented step by step in the sections below.

## Prerequisites

Firstly, we load the required libraries (you need to install them if you don't
have them):
```{r Load libraries, message = FALSE}
library(rje) # version 1.9
library(igraph) # version 1.2.2, see (Csardi, 2006)
library(superheat) # version 1.0.0, see (Barter, 2017)
```
and the relevant helper functions:
```{r Helper functions}
# Set the working directory to the gitsbe-model-analysis folder: 
# setwd("pathTo/gitsbe-model-analysis")
source("Rscripts/input_functions.R")
source("Rscripts/output_functions.R")
source("Rscripts/analysis_functions.R")
source("Rscripts/plot_functions.R")
```

The R version used for this analysis is: 
```{r R version, results = 'asis'}
pretty.print.string(R.version.string)
```

## Input

We will define the name of the cell line which must match the name of the 
directory that has the input files inside the `cell-lines-2500` directory. Our
analysis in this notebook will be done on the data for the `MDA-MB-468` cell line:
```{r Input: cell line}
cell.line = "MDA-MB-468"
data.dir = paste0(getwd(), "/cell-lines-2500/", cell.line, "/")
```
**Three inputs** are used in this analysis:

- The **model_predictions** file which has for each model the prediction for 
each drug combination tested (*0* = no synergy predicted, *1* = synergy 
predicted, *NA* = couldn't find stable states in either the drug combination 
inhibited model or in any of the two single-drug inhibited models)
- The **observed_synergies** file which lists the drug combinations that were 
observed as synergistic for the particular cell line.
- The **models** directory, which is the same as the models directory produced 
by `Gitsbe` and has one `.gitsbe` file per model that includes this info:
    - The *stable state* of the boolean model. Note that a model can have 1 
stable state or none in our simulations - but the models used in this analysis 
have been selected through a genetic evolution algorithm in `Gitsbe` and so in 
the end, only those with 1 stable state have higher fitness values and remain in 
the later generations. Higher fitness here means a better match of a model's 
stable state to the cell line derived steady state (a perfect match would result 
in a fitness of 1)
    - The *boolean equations* of the model

```{r Input: files}
model.predictions.file = paste0(data.dir, "model_predictions")
observed.synergies.file = paste0(data.dir, "observed_synergies")
models.dir = paste0(data.dir, "models")
```
Now, we parse the data into proper R objects. First the synergy predictions 
per model:
```{r Input: model predictions}
model.predictions = get.model.predictions(model.predictions.file)
head(model.predictions)
```
So, we can see that our dataset has the models as rows and each column is a 
different drug combination that was tested in our simulations.
```{r Model&Drug Stats, results = 'asis'}
drug.combinations.tested = colnames(model.predictions)
models                   = rownames(model.predictions)
nodes                    = get.node.names(models.dir)

number.of.drug.comb.tested = length(drug.combinations.tested)
number.of.models           = length(models)
number.of.nodes            = length(nodes)

print.model.and.drug.stats(number.of.drug.comb.tested, number.of.models, 
                           number.of.nodes, html.output = TRUE)
```

Next, we get the full stable state and the equations per model:
```{r Input: models stable states}
models.stable.state = get.stable.state.from.models.dir(models.dir)
models.stable.state = models.stable.state[models,]
head(as.data.frame(models.stable.state))
```
The rows of the above dataset represent the models while the columns are the
names of the nodes (proteins, genes, etc.) of the cancer cell network under
study. So, each model has one stable state which means that in every model, the
nodes in the network have reached a state of either 0 (inhibition) or 1
(activation).
```{r Input: models equations}
models.equations = get.equations.from.models.dir(
  models.dir, remove.equations.without.link.operator = TRUE)
models.equations = models.equations[models,]
head(as.data.frame(models.equations))
```
For the equations, if we look at a specific row (a model so to speak), the 
columns (node names) correspond to the *targets of regulation* (and the network 
has been built so that *every node is a target* - i.e. it has other nodes 
activating and/or inhibiting it). The general form of a boolean equation is: 
**Target \*= (Activator OR Activator OR...) AND NOT (Inhibitor OR Inhibitor 
OR...)**.

The difference between the models' boolean equations is the *link operator* 
(OR NOT/AND NOT) which has been mutated (changed) through the evolutionary 
process of the genetic algorithm in `Gitsbe`. For example, if a model has for 
the column `ERK_f` (of the `models.equations` object) a value of 1, the correspoding 
equation is: ERK_f \*= (MEK_f) **OR NOT** ((DUSP6) OR PPP1CA). A value of 0 would 
correspond to the same equation but having **AND NOT** as the link operator:
RK_f \*= (MEK_f) **AND NOT** ((DUSP6) OR PPP1CA). Note that the equations that do 
not have link operators (meaning that they are *the same for every model*) are 
discarded (so less columns in this dataset) since in a later section we study only 
the equations whose link operators differentiate between the models.

Lastly, the synergies observed for this particular cell line are:
```{r Input: observed synergies, results = 'asis'}
observed.synergies = get.observed.synergies(
  observed.synergies.file, drug.combinations.tested)
number.of.observed.synergies = length(observed.synergies)

pretty.print.vector.values(observed.synergies, 
                           vector.values.str = "observed synergies")
```

## Analysis

### Performance Statistics

It will be interesting to know the percentage of the above observed synergies 
that were actually predicted by at least one of the models (there might be 
combinations that no model in our dataset could predict):
```{r Synergy stats 1}
# Subset the model.data to the observed synergies
observed.model.predictions = 
  model.predictions[,sapply(drug.combinations.tested, function(drug.comb) {
    is.correct.synergy(drug.comb, observed.synergies)
  })]

# Subset the model.data to the unobserved synergies
unobserved.model.predictions = 
  model.predictions[,sapply(drug.combinations.tested, function(drug.comb) {
    !is.correct.synergy(drug.comb, observed.synergies)
  })]

stopifnot(dim(observed.model.predictions)[2] + 
          dim(unobserved.model.predictions)[2] == number.of.drug.comb.tested)

number.of.models.per.observed.synergy = 
  apply(observed.model.predictions, 2, sum, na.rm = T)
predicted.synergies = names(number.of.models.per.observed.synergy)[
                            number.of.models.per.observed.synergy > 0]
```

```{r Print synergy stats, results = 'asis'}
pretty.print.vector.values(predicted.synergies, 
                           vector.values.str = "predicted synergies")

predicted.synergies.percentage = 100 * length(predicted.synergies) /
                    number.of.observed.synergies
pretty.print.string(paste0("Percentage of True Positive predicted synergies: ", 
                    specify.decimal(predicted.synergies.percentage, 2), "%"))
```

So, for this particular cell line, **there were indeed observed synergies that no 
model could predict** (e.g. `AK-BI`, `AK-PI`). Next, we would like to know the maximum 
number of observed synergies predicted by one model alone - **can one model by itself 
predict all the true positive synergies predicted by all the models together** or 
do we need many models to capture this diverse synergy landscape? To do that, we 
go even further and **count the number of models that predict a specific set of 
observed synergies** for every possible combination subset of the `predicted.synergies`
object:
```{r Synergy stats 2}
# Find the number of predictive models for every synergy subset
predicted.synergies.powerset = powerSet(predicted.synergies)
predicted.synergies.powerset = predicted.synergies.powerset[
  order(sapply(predicted.synergies.powerset, length))
]
names(predicted.synergies.powerset) = 
  sapply(predicted.synergies.powerset, function(drug.comb.set) {
    paste(drug.comb.set, collapse = ",")
  }) 
synergy.subset.stats = 
  sapply(predicted.synergies.powerset, function(drug.comb.set) {
    count.models.that.predict.synergy.set(drug.comb.set, observed.model.predictions) 
})

# Bar plot of the number of models for every possible observed synergy combination set
# Tweak the threshold.for.subset.removal and bottom.margin as desired
make.barplot.on.synergy.subset.stats(synergy.subset.stats,
                                     threshold.for.subset.removal = 1, 
                                     bottom.margin = 9, cell.line)
```

From the above figure (where we excluded sets of synergies that were predicted 
by no model by setting the `threshold.for.subset.removal` value to 1) we observe 
that:

- **More than half of the models predict no synergies**
- The `PI-D1` synergy is the most common predicted synergy for the rest of the models

Next we calculate the maximum number of correctly predicted synergies ($TP$ - 
True Positives) per model:
```{r TP Stats}
# Count the predictions of the observed synergies per model (TP)
models.synergies.tp = apply(observed.model.predictions, 1, sum, na.rm = T)
models.synergies.tp.stats = table(models.synergies.tp)

# Bar plot of number of models vs correctly predicted synergies
make.barplot.on.models.stats(models.synergies.tp.stats, cell.line, 
                            title = "True Positive Synergy Predictions",
                            xlab = "Number of maximum correctly predicted synergies",
                            ylab = "Number of models")
```

To summarize:

- There were **59 models that predicted 3 synergies** - the set 
`PI-JN,PI-D1,JN-D1` - which is the **maximum number of predicted synergies by an 
individual model**
- **No model could predict all 5 of the total predicted synergies**

The power of the ensemble model approach lies in the fact that (as 
we saw from the above figures) even though we may not have *individual super models* 
that can predict many observed drug combinations, there are many that predict 
*at least one* and which will be used by the drug response analysis module 
(`Drabme`) to better infer the synergistic drug combinations. It goes without 
saying though, that the existance of models that could predict more than a 
handful of synergies would be beneficial for any approach that performs drug 
response analysis on a multitude of models.

### Biomarker analysis

#### Intro-Methods

Now, we want to investigate and find possible important nodes - **biomarkers** -
whose activity state either distinguishes good performance models from less 
performant ones (in terms of a performance metric - e.g. the true positive 
synergies predicted) or makes some models predict a specific synergy
compared to others that can't (but could predict other synergies). So, we devised 
**two strategies** to split the models in our disposal to *good* and *bad* ones 
(but not necessarily all of them), the demarcation line being either a performance 
metric (the number of $TP$ or the Matthews Correlation Coefficient score) or 
the prediction or not of a specific synergy. Then, for 
each group of models (labeled as either *good* or *bad*) we find the average 
activity state of every node in the network (value between 0 and 1) and then we 
compute the average state difference for each node between the two groups:
$\forall i\in nodes,mean(state_i)_{good} - mean(state_i)_{bad}$. 
Our hypothesis is that if the absolute value of these average differences are 
larger than a **user-defined threshold (e.g. 0.7)** for a small number of nodes 
(the less the better) while for the rest of the nodes they remain close to zero, 
then the former nodes are considered **the most important** since they 
define the difference between the *average bad model* and the *average good one*
in that particular case study. We will also deploy a network visualization method 
to observe these average differences.

#### True Positives-based analysis

Using our first strategy, **we will split the models based on the number of true 
positive predictions**. For example, the bad models will be the ones that 
predicted 0 $TP$ synergies whereas the good models will be the ones that predicted 
2 $TP$ (we will denote the grouping as $(0,2)$). This particular classification 
strategy will be used for every possible combination of the number of $TP$ as 
given by the `models.synergies.tp.stats` object and the density estimation of the
nodes' average state differences in each case will be ploted in a common graph:
```{r Good vs Bad models based on the number of TPs}
tp.values = as.numeric(names(models.synergies.tp.stats))
tp.values.comb = t(combn(tp.values, 2))

diff.tp.results = apply(tp.values.comb, 1, function(comb) {
  return(get.avg.activity.diff.based.on.tp.predictions(
    models, models.synergies.tp, models.stable.state, 
    num.low = comb[1], num.high = comb[2]))
})

tp.comb.names = apply(tp.values.comb, 1, function(row) {
  return(paste0("(", paste(row, collapse = ","), ")")) 
})
colnames(diff.tp.results) = tp.comb.names
diff.tp.results = t(diff.tp.results)

tp.densities = apply(abs(diff.tp.results), 1, density)
make.multiple.density.plot(tp.densities, legend.title = "True Positives", 
                           x.axis.label = "Activity state (absolute difference value)",
                           title = "Density Estimation for the Average State Difference")
```

What we are actually looking for is density plots that are *largely skewed to 
the right* (so the average absolute differences of these nodes are close to zero)
while there are a few areas of non-zero densities which are as close to 1 as 
possible. So, from the above graph, the density plots that fit this description 
are the ones marked as $(1,3)$ and $(2,3)$. We will visualize the nodes'
average state differences in a network graph, where the color of each node will 
denote how much more inhibited or active that node is, in the average good model 
vs the average bad one. The color of the edges will denote activation (green) or 
inhibition (red). We first build the network from the node topology (edge list):
```{r Input: topology}
parent.dir = get.parent.dir(data.dir)
topology.file = paste0(parent.dir, "/topology")
coordinates.file = paste0(parent.dir, "/network_xy_coordinates")

net = contruct.network(topology.file, models.dir)

# a static layout for plotting the same network always (igraph)
# nice.layout = layout_nicely(net)
nice.layout = as.matrix(read.table(coordinates.file))
```
In the next colored graphs we can identify the important nodes whose activity
state can influence the true positive prediction performance (from 0 true 
positive synergies to a total of 3):
```{r Graphs: Good vs Bad models based on the number of TPs}
# useful tutorial about network visualization in R: (Ognyanova, 2018)

diff.tp.0.3 = diff.tp.results[3,]
plot.network(net, diff.tp.0.3, layout = nice.layout,
             title = "Bad models (0 TP) vs Good models (3 TP)")

diff.tp.1.3 = diff.tp.results[5,]
plot.network(net, diff.tp.1.3, layout = nice.layout,
             title = "Bad models (1 TP) vs Good models (3 TP)")

diff.tp.2.3 = diff.tp.results[6,]
plot.network(net, diff.tp.2.3, layout = nice.layout,
             title = "Bad models (2 TP) vs Good models (3 TP)")
```

We will now **list the important nodes** that affect the models' prediction 
performance **with regards to the number of true positive synergies** that they 
predict. Comparing the graphs above, we observe that there exist common nodes 
that maintain the same significant influence in all of the graphs. We set the 
**threshold for the absolute significance level in average state differences 
to $0.7$**. A node will be marked as a biomarker (active or inhibited) if 
its activity state difference surpassed the aforementioned threshold (positively 
or negatively) **for any of the tested groups** (e.g. 2 $TP$ vs 3 $TP$). So, the 
nodes that have to be in a more *active* state are:
```{r TP biomarkers (active), results = 'asis'}
biomarkers.tp.active = get.biomarkers(diff.tp.results, threshold = 0.7, 
                                      type = "active")
pretty.print.vector.values(biomarkers.tp.active)
```

Also, the nodes that have to be in a more *inhibited* state are:
```{r TP biomarkers (inhibited), results = 'asis'}
biomarkers.tp.inhibited = get.biomarkers(diff.tp.results, threshold = 0.7, 
                                         type = "inhibited")
pretty.print.vector.values(biomarkers.tp.inhibited)
```

#### MCC classification-based analysis

The previous method to split the models based on the number of true positive 
predictions is a good metric for absolute performance but a very restricted one 
since it ignores the other values of the confusion matrix for each model (true 
negatives, false positives, false negatives). Also, since our dataset is 
*imbalanced* in the sense that out of the total drug combinations tested only a 
few of them are observed as synergistic (and in a hypothetical larger drug 
screening evaluation it will be even less true positives) we will now devise a 
method to split the models into different performance categories based on the 
value of the **Matthews Correlation Coefficient (MCC) score** which takes into 
account the balance ratios of all the four confusion matrix values:
```{r MCC Stats}
# Count the false negatives (FN)
models.synergies.fn = apply(observed.model.predictions, 1, function(x) {
  sum( x == 0 | is.na(x) )
})

# P = TP + FN (Positives)
positives = ncol(observed.model.predictions)
models.synergies.p = models.synergies.tp + models.synergies.fn

# Count the predictions of the non-observed synergies per model (FP)
models.synergies.fp = apply(unobserved.model.predictions, 1, sum, na.rm = T)

# Count the True Negatives (TN)
models.synergies.tn = apply(unobserved.model.predictions, 1, function(x) {
  sum( x == 0 | is.na(x))
})

# N = FP + TN (Negatives)
negatives = ncol(unobserved.model.predictions)
models.synergies.n = models.synergies.fp + models.synergies.tn

# checks
stopifnot(models.synergies.p == positives)
stopifnot(models.synergies.n == negatives)
stopifnot(positives + negatives == number.of.drug.comb.tested)

# Calculate Matthews Correlation Coefficient (MCC)
models.mcc = calculate.mcc(models.synergies.tp, models.synergies.tn, 
                           models.synergies.fp, models.synergies.fn,
                           positives, negatives)
models.mcc.stats = table(models.mcc, useNA = "ifany")

make.barplot.on.models.stats(models.mcc.stats, cell.line, title = "MCC scores", 
                             xlab = "MCC value", ylab = "Number of models", 
                             cont.values = TRUE, threshold = 10)
```

Note that for presentation purposes in the figure above, we pruned some MCC-bars 
with lower model frequency values. We observe that:

- There are no relatively bad models (MCC values close to -1)
- Some models perform close to random prediction ($MCC\approx0$)
- There are models that had `NaN` value for the MCC score

Given the MCC formula:
$MCC = (TP\cdot TN - FP\cdot FN)/\sqrt{(TP+FP) * (TP+FN) * (TN+FP) * (TN+FN)}$,
we can see that the MCC value can be `NaN` because of zero devision. Two of the 
four values in the denominator represent the number of positive $(TP+FN)$ and 
negative $(TN+FP)$ observations which are non-zero for every model, since they 
correspond to the observed and non-obsered synergies in each case. 
The case where both $TN$ and $FN$ are zero is rare (if non-existent) 
because of the imbalanced dataset (large proportion of negatives) and the reason 
that logical models which report no negatives means that they should find 
fixpoint attractors for every possible drug combination perturbation which also 
is extremely unlikely. We can actually see that the `NaN` are produced by models 
that have **both TP and FP equal to zero**:
```{r NaN MCC score, results = 'asis'}
pretty.print.string(sum(models.synergies.tp + models.synergies.fp == 0))
```

Since these models could intentify no synergies (either correctly or 
wrongly), we decided to put them as the lowest performant category in our 
MCC-based analysis. To classify the models based on the MCC score (which takes 
values in the $[-1, 1]$ interval), we devised a method that **splits the previously 
found MCC values range to intervals of a specific size** (each corresponding to a 
different classification category), so that each model's MCC score falls within 
a specific interval:
```{r MCC classification, results = 'asis'}
mcc.values = as.numeric(names(models.mcc.stats))

# split into classes with a 0.2 MCC value range each
mcc.intervals = get.mcc.intervals(mcc.values, interval.size = 0.2)

# add NaN category (if applicable)
if (sum(is.na(mcc.values)) > 0) {
  mcc.intervals = rbind(c(NaN, NaN), mcc.intervals)
}

mcc.classes = get.mcc.classes(mcc.intervals)
print.mcc.classification.info(mcc.classes, html.output = TRUE)
```

Following our first strategy, **we will split the models based on the MCC 
performance metric score**. For example, the bad models will be the ones that 
had a `NaN` MCC score $(TP+FP = 0)$ whereas the good models will be the ones that 
had an MCC score between $[-0.2, 0)$. This particular classification strategy will 
be used for every possible combination of the MCC classes as defined by the 
`mcc.classes` object and the density estimation of the nodes' average state 
differences in each case will be ploted in a common graph:
```{r Good vs Bad models based on the MCC classification}
mcc.class.id.comb = t(combn(1:length(mcc.classes), 2))

diff.mcc.results = apply(mcc.class.id.comb, 1, function(comb) {
  return(get.avg.activity.diff.based.on.mcc.classification(
    models, models.mcc, mcc.intervals, models.stable.state, 
    class.id.low = comb[1], class.id.high = comb[2]))
})

mcc.classes.comb.names = apply(mcc.class.id.comb, 1, function(comb) {
  return(paste0("(", mcc.classes[comb[1]], ", ", mcc.classes[comb[2]], ")"))
})

colnames(diff.mcc.results) = mcc.classes.comb.names
diff.mcc.results = t(diff.mcc.results)

mcc.densities = apply(abs(diff.mcc.results), 1, density)
make.multiple.density.plot(mcc.densities, legend.title = "MCC classes",
                           x.axis.label = "Activity state (absolute difference value)",
                           title = "Density Estimation for the Average State Difference")
```

The density plots of interest here are the ones that compare the **1st vs 4th class** 
$(NaN, [0.2, 0.4])$ and the **2nd vs 4th class** $([-0.2, 0), [0.2, 0.4])$. Next, 
we visualize the average state differences with our network coloring method, in 
order to identify the important nodes whose activity state can influence the 
prediction performance based on the MCC classification:
```{r Graphs: Good vs Bad models based on the MCC classification}
# (NaN, [0.2, 0.4])
diff.mcc.1.4 = diff.mcc.results[3,]
plot.network(net, diff.mcc.1.4, layout = nice.layout, 
             title = "Bad models, MCC: NaN vs Good models, MCC: [0.2, 0.4]")

# ([-0.2, 0), [0.2, 0.4])
diff.mcc.2.4 = diff.mcc.results[5,]
plot.network(net, diff.mcc.2.4, layout = nice.layout, 
             title = "Bad models, MCC: [-0.2, 0) vs Good models, MCC: [0.2, 0.4]")
```

We will now **list the important nodes** that affect the models' prediction 
performance **with regards to their MCC classification**, using the same method
as in the `True Positives-based analysis` section. First, the nodes that have to 
be in a more *active* state:
```{r MCC biomarkers (active), results = 'asis'}
biomarkers.mcc.active = get.biomarkers(diff.mcc.results, threshold = 0.7, 
                                      type = "active")
pretty.print.vector.values(biomarkers.mcc.active)
```

Secondly, the nodes that have to be in a more *inhibited* state:
```{r MCC biomarkers (inhibited), results = 'asis'}
biomarkers.mcc.inhibited = get.biomarkers(diff.mcc.results, threshold = 0.7, 
                                      type = "inhibited")
pretty.print.vector.values(biomarkers.mcc.inhibited)
```

Compared to the true positives-based classification, the MCC classification finds
**no inhibited biomarkers** at the $0.7$ threshold.

#### Common performance-related biomarkers

Comparing the two methods we used for the classification of the models' prediction
performance (TP and MCC), we observe that there exist **no common biomarkers** in 
either active or inhibited state cases:
```{r Common biomarkers (mixed), results = 'asis'}
common.biomarkers.mixed.1 = 
  get.common.values(biomarkers.mcc.active, biomarkers.tp.inhibited)
common.biomarkers.mixed.2 = 
  get.common.values(biomarkers.mcc.inhibited, biomarkers.tp.active)
common.biomarkers.active = 
  get.common.values(biomarkers.mcc.active, biomarkers.tp.active)
common.biomarkers.inhibited =
  get.common.values(biomarkers.mcc.inhibited, biomarkers.tp.inhibited)
```

We believe that the MCC score better captures the performance categories within 
our imbalanced dataset so we will use the results from the MCC analysis as the 
performance biomarkers. Lastly, we will save these performance-related biomarkers 
for further analysis and comparison with the results from all the different cell 
lines:
```{r Save performance biomarkers}
biomarkers.dir = paste0(data.dir, "biomarkers/")
save.vector.to.file(vector = biomarkers.mcc.active, 
                    file = paste0(biomarkers.dir, "biomarkers_active"))
save.vector.to.file(vector = biomarkers.mcc.inhibited, 
                    file = paste0(biomarkers.dir, "biomarkers_inhibited"))
```

#### Equation-based analysis

It will be interesting to see the different patterns in the form of the boolean 
equations (regarding the mutation of the *link operator* as mentioned in the `Input` 
section) when comparing higher performance models vs the low performant ones. We 
could also check if any of the biomarkers found above relate to a different link 
operator on average between models with different performance characteristics 
(e.g. higher predictive models should have the `OR NOT` as the link operator in 
a boolean equation where a specific biomarker is the regulation target) or if they 
constitute targets of exclusively activating nodes or inhibiting ones (an equation
with no link operator).

The performance metric we will first use to sort the models is the number of true 
positive predictions:
```{r True Positives Coloring}
number.of.unique.tp.values = length(tp.values)
rbPal = colorRampPalette(c("red", "black", "green"))
tp.colors = rbPal(number.of.unique.tp.values)
models.synergies.tp.colors = tp.colors[
  as.numeric(cut(models.synergies.tp, breaks = number.of.unique.tp.values))
]
names(models.synergies.tp.colors) = models

title = "Color Palette"
xlab  = "Number of True Positives"
make.color.bar.plot(color.vector = tp.colors, 
                    number.vector = tp.values, 
                    title, xlab)
```

So, we will now illustrate the heatmap of the `models.equations` object raw-order 
by the number of $TP$ predictions:
```{r Equation Heatmap (TP raw-ordered), fig.height = 7, fig.width = 7, dpi = 300}
# order based on number of true positives
models.synergies.tp.sorted = sort(models.synergies.tp)
models.sorted = names(models.synergies.tp.sorted)

models.equations.sorted = models.equations[models.sorted, ]
models.synergies.tp.colors.sorted = models.synergies.tp.colors[models.sorted]

# color biomarkers names in the heatmap
bottom.nodes.colors = rep("black", length(colnames(models.equations.sorted)))
names(bottom.nodes.colors) = colnames(models.equations.sorted)
bottom.nodes.colors[names(bottom.nodes.colors) %in% 
                    biomarkers.mcc.active] = "green"
bottom.nodes.colors[names(bottom.nodes.colors) %in% 
                    biomarkers.mcc.inhibited] = "red"

superheat(models.equations.sorted, title.alignment = "center",
          title = "Models link operators (AND/OR NOT)", 
          row.title = "Models (TP sorted)", smooth.heat = TRUE,
          row.dendrogram = FALSE, col.dendrogram = TRUE, scale = FALSE,
          clustering.method = "hierarchical", dist.method = "euclidean",
          force.left.label = TRUE, left.label.text.size = 0, 
          left.label.size = 0.05, 
          left.label.col = models.synergies.tp.colors.sorted,
          heat.pal = c("red","gold"),
          legend.breaks = c(0, 1), 
          grid.hline = FALSE, grid.vline = FALSE,
          bottom.label.text.size = 3, bottom.label.text.angle = 90, 
          bottom.label.col = "white", bottom.label.text.col = bottom.nodes.colors)
```

In the heatmap above, an equation whose link operator is `AND NOT` is represented 
with red color while an `OR NOT` link operator is represented with gold color.
The targets whose equations do not have a link operator are not represented.
The rows are ordered by the number of true positive predictions (ascending). 
We have colored the names of the network nodes that were also found as biomarkers 
(green color is used for the active biomarkers and red color for the inhibited 
biomarkers). We observe that:

- **Most of the biomarkers are nodes that do not have both activators and inhibitors**
and so are absent from the above heatmap
- There doesn't seem to exist a pattern between the models' link operators 
and their corresponding performance (at least not for all of the nodes) when using
the true positive predictions as a classifier for the models
- There exist a lot of target nodes that need to have the `OR NOT` link operator 
in their respective boolean equation in order for the corresponding logical model 
to show a higher number of true positive predictions. By assigning the `OR NOT` 
link operator to a target's boolean regulation equation, we allow more flexibility 
to the target's output active result state - meaning that the inhibitors play less 
role and the output state has a higher probability of being active - compared to 
assigning the `AND NOT` link operator to the equation

We will also illustrate the heatmap of the `models.equations` object raw-order 
by the MCC score which is a better performance classifier. Models who had a `NaN` 
MCC score will be again placed in the lower performant category:
```{r Equation Heatmap (MCC raw-ordered), fig.height = 7, fig.width = 7, dpi = 300, warning = FALSE}
# order based on the MCC value
models.mcc.sorted = sort(models.mcc, na.last = FALSE)
models.sorted = names(models.mcc.sorted)

models.equations.sorted = models.equations[models.sorted, ]

superheat(models.equations.sorted, title.alignment = "center",
          title = "Models link operators (AND/OR NOT)", 
          row.title = "Models (MCC sorted)", smooth.heat = TRUE,
          row.dendrogram = FALSE, col.dendrogram = TRUE, scale = FALSE,
          clustering.method = "hierarchical", dist.method = "euclidean",
          yr = models.mcc.sorted, yr.plot.type = "line", yr.axis.name = "MCC score",
          yr.line.col = "cornflowerblue", yr.plot.size = 0.2,
          heat.pal = c("red","gold"),
          legend.breaks = c(0, 1), 
          grid.hline = FALSE, grid.vline = FALSE,
          bottom.label.text.size = 3, bottom.label.text.angle = 90, 
          bottom.label.col = "white", bottom.label.text.col = bottom.nodes.colors)
```

In the MCC-ordered heatmap above we observe that there is **better correlation 
between the boolean equations' link operators and the performance of a model** 
compared to the $TP$ classification. For example, the node `STAT3` and the biomarker 
`RAF1` show distinguished patterns for the higher performance models (use of 
`AND NOT` and `OR NOT` link operators respectively) when the models are sorted 
by the MCC score.

#### Synergy-prediction based analysis

We will now use the second strategy to split the models, based on whether 
they predict a specific observed synergy or not. This will allow us
to **find biomarkers that affect the prediction of a specific synergy**. 
For example, the good models could be the ones that predicted the hypothetical
synergy `A-B` while the bad models all the rest that identified the particular 
combination as non-synergistic.
In another case scenario, the good models could be those that predicted a 
triple synergy set `A-B,C-D,A-C`, while the bad models could be the ones 
that predicted the double synergy subset `A-B,C-D` (excluding the common models
that predicted both the triple synergy set and subsequently its given subset). 
In such a case scenario, we want to find out which nodes are responsible for making 
the good models **predict the extra synergy** - in this hypothetical case the 
synergy `A-C` - demonstrating thus better model performance. Note that the models 
selected in each case as good or bad, could have predicted other synergies as well 
(correctly as $TP$ or wrongly as $FP$) which means that the biomarker selection 
method could be somewhat innacurate, since we can't really know the prediction of 
which extra synergy or synergies the biomarkers' state affected. To account for 
this, we label as good models those that predict large synergy sets (so fewer 
models) which capture almost all the true positive predictions and also minimize 
the possible extra different synergies predicted by models of the same classification 
category (e.g. the good models).

Starting with the first model classification method (prediction vs non-prediction 
of a particular synergy), we generate the density distribution of the nodes' 
average state differences between the good and bad models for each predicted synergy:
```{r Good vs Bad models based on a specific synergy prediction}
diff.predicted.synergies.results = 
  sapply(predicted.synergies, function(drug.comb) {
    get.avg.activity.diff.based.on.specific.synergy.prediction(
      model.predictions, models.stable.state, drug.comb)
  })
diff.predicted.synergies.results = t(diff.predicted.synergies.results)

densities = apply(abs(diff.predicted.synergies.results), 1, density)
make.multiple.density.plot(densities, legend.title = "Predicted Synergies",
                           x.axis.label = "Activity state (absolute difference value)",
                           title = "Density Estimation for the Average State Difference")
```

Next, we will plot the biomarkers with the network visualization method (for all
predicted synergies) and output the respective biomarkers in each case. 
The biomarker results for each predicted synergy will be stored for further 
comparison with the results from the other cell lines.

```{r Graphs: Good vs Bad models based on a specific synergy prediction}
threshold = 0.7
for (drug.comb in predicted.synergies) {
  title.text = paste0("Prediction of ", drug.comb, 
                      " synergy: Good models vs Bad Models")
  diff = diff.predicted.synergies.results[drug.comb, ]
  plot.network(net, diff, layout = nice.layout, title = title.text)
  
  biomarkers.active = diff[diff > threshold]
  biomarkers.inhibited = diff[diff < -threshold]
  
  save.vector.to.file(vector = biomarkers.active, file = paste0(
    biomarkers.dir, drug.comb, "_biomarkers_active"), with.row.names = TRUE)
  save.vector.to.file(vector = biomarkers.inhibited, file = paste0(
    biomarkers.dir, drug.comb, "_biomarkers_inhibited"), with.row.names = TRUE)
}
```

```{r Biomarkers per predicted synergy, results = 'asis'}
print.biomarkers.per.predicted.synergy(biomarkers.dir, drug.comb,
                                       predicted.synergies)
```

We notice that for some predicted synergies the above method identified zero 
biomarkers. We will now study cases where the main goal is the better identification 
and/or refinement of the biomarkers responsible for allowing the models to 
predict one extra synergy from a specific synergy set. We will focus on the 
predicted synergies `PI-JN`, `JN-D1`, `D1-P5` and `PI-D1`.

#### Synergy-set prediction based analysis

##### PI-JN synergy

The first use case will contrast the models that predicted the synergy set 
`PI-JN,JN-D1` vs the models that predicted the single synergy `JN-D1`. 
This will allow us to find important nodes whose state can influence the 
prediction performance of a model and make it predict the extra `PI-JN` synergy:
```{r Good vs Bad models based on different synergy set prediction (PI-JN)}
synergy.set.str    = "PI-JN,JN-D1"
synergy.subset.str = "JN-D1"
diff.PI.JN = get.avg.activity.diff.based.on.diff.synergy.set.prediction(
  synergy.set.str, synergy.subset.str, model.predictions, models.stable.state)
```
We now visualize the average state differences:
```{r Graphs: Good vs Bad models based on different synergy set prediction (PI-JN)}
title.text = paste0("Good models (", synergy.set.str,
                    ") vs Bad Models (", synergy.subset.str, ")")
plot.network(net, diff.PI.JN, layout = nice.layout, title = title.text)
```

and report the active biomarkers found:
```{r PI-JN biomarkers (active), results = 'asis'}
biomarkers.PI.JN.active = diff.PI.JN[diff.PI.JN > threshold]
pretty.print.vector.names(biomarkers.PI.JN.active)
```

The inhibited biomarkers are:
```{r PI-JN biomarkers (inhibited), results = 'asis'}
biomarkers.PI.JN.inhibited = diff.PI.JN[diff.PI.JN < -threshold]
pretty.print.vector.names(biomarkers.PI.JN.inhibited)
```

Lastly, we add the newly found biomarkers in the respective files:
```{r Update PI-JN biomarkers}
drug.comb = "PI-JN"
update.biomarkers.files(biomarkers.dir, drug.comb, 
                        biomarkers.PI.JN.active, biomarkers.PI.JN.inhibited)
```

##### JN-D1 synergy

The second use case will contrast the models that predicted the synergy set 
`PI-JN,JN-D1` vs the models that predicted the single synergy `PI-JN`. 
This will allow us to find important nodes whose state can influence 
the prediction performance of a model and make it predict the extra `JN-D1` synergy:
```{r Good vs Bad models based on different synergy set prediction (JN-D1)}
synergy.set.str    = "PI-JN,JN-D1"
synergy.subset.str = "PI-JN"
diff.JN.D1 = get.avg.activity.diff.based.on.diff.synergy.set.prediction(
  synergy.set.str, synergy.subset.str, model.predictions, models.stable.state)
```
We now visualize the average state differences:
```{r Graphs: Good vs Bad models based on different synergy set prediction (JN-D1)}
title.text = paste0("Good models (", synergy.set.str,
                    ") vs Bad Models (", synergy.subset.str, ")")
plot.network(net, diff.JN.D1, layout = nice.layout, title = title.text)
```

and report the active biomarkers found:
```{r JN-D1 biomarkers (active), results = 'asis'}
biomarkers.JN.D1.active = diff.JN.D1[diff.JN.D1 > threshold]
pretty.print.vector.names(biomarkers.JN.D1.active)
```

The inhibited biomarkers are:
```{r JN-D1 biomarkers (inhibited), results = 'asis'}
biomarkers.JN.D1.inhibited = diff.JN.D1[diff.JN.D1 < -threshold]
pretty.print.vector.names(biomarkers.JN.D1.inhibited)
```

Lastly, we add the newly found biomarkers in the respective files:
```{r Update JN-D1 biomarkers}
drug.comb = "JN-D1"
update.biomarkers.files(biomarkers.dir, drug.comb, 
                        biomarkers.JN.D1.active, biomarkers.JN.D1.inhibited)
```

##### D1-P5 synergy

The third use case will contrast the models that predicted the synergy set 
`PI-D1,D1-P5` vs the models that predicted the single synergy `PI-D1`. This will
allow us to find important nodes whose state can influence the prediction 
performance of a model and make it predict the extra `D1-P5` synergy:
```{r Good vs Bad models based on different synergy set prediction (D1-P5)}
synergy.set.str = "PI-D1,D1-P5"
synergy.subset.str = "PI-D1"
diff.D1.P5 = get.avg.activity.diff.based.on.diff.synergy.set.prediction(
  synergy.set.str, synergy.subset.str, model.predictions, models.stable.state)
```
We now visualize the average state differences:
```{r Graphs: Good vs Bad models based on different synergy set prediction (D1-P5)}
title.text = paste0("Good models (", synergy.set.str,
                    ") vs Bad Models (", synergy.subset.str, ")")
plot.network(net, diff.D1.P5, layout = nice.layout, title = title.text)
```

Sadly, there were no biomarkers found at the $0.7$ threshold in either active or
inhibited states:
```{r D1-P5 biomarkers (mixed), results = 'asis'}
biomarkers.D1.P5.active = diff.D1.P5[diff.D1.P5 > threshold]
pretty.print.vector.names(biomarkers.D1.P5.active)

biomarkers.D1.P5.inhibited = diff.D1.P5[diff.D1.P5 < -threshold]
pretty.print.vector.names(biomarkers.D1.P5.inhibited)
```

##### PI-D1 synergy

The fourth use case will contrast the models that predicted the synergy set 
`PI-D1,D1-P5` vs the models that predicted the single synergy `D1-P5`. This will
allow us to find important nodes whose state can influence the prediction 
performance of a model and make it predict the extra `PI-D1` synergy:
```{r Good vs Bad models based on different synergy set prediction (PI-D1)}
synergy.set.str = "PI-D1,D1-P5"
synergy.subset.str = "D1-P5"
diff.PI.D1 = get.avg.activity.diff.based.on.diff.synergy.set.prediction(
  synergy.set.str, synergy.subset.str, model.predictions, models.stable.state)
```
We now visualize the average state differences:
```{r Graphs: Good vs Bad models based on different synergy set prediction (PI-D1)}
title.text = paste0("Good models (", synergy.set.str,
                    ") vs Bad Models (", synergy.subset.str, ")")
plot.network(net, diff.PI.D1, layout = nice.layout, title = title.text)
```

and report the active biomarkers found:
```{r PI-D1 biomarkers (active), results = 'asis'}
biomarkers.PI.D1.active = diff.PI.D1[diff.PI.D1 > threshold]
pretty.print.vector.names(biomarkers.PI.D1.active)
```

The inhibited biomarkers are:
```{r PI-D1 biomarkers (inhibited), results = 'asis'}
biomarkers.PI.D1.inhibited = diff.PI.D1[diff.PI.D1 < -threshold]
pretty.print.vector.names(biomarkers.PI.D1.inhibited)
```

Lastly, we add the newly found biomarkers in the respective files:
```{r Update PI-D1 biomarkers}
drug.comb = "PI-D1"
update.biomarkers.files(biomarkers.dir, drug.comb, 
                        biomarkers.PI.D1.active, biomarkers.PI.D1.inhibited)
```

#### Biomarker results

In this section, we will compare the **biomarkers found per predicted synergy** 
for this particular cell line as well as the **performance biomarkers** (notated 
as **PERF** in the heatmap below) which are based on the results from the MCC 
classification-based model analysis:
```{r Biomarkers heatmap, fig.height = 5, fig.width = 7, dpi = 300}
# Biomarkers from sections:
# `Synergy-prediction based analysis`, `Synergy-set prediction based analysis`
biomarkers.synergy.res =
  get.biomarkers.per.synergy(predicted.synergies, biomarkers.dir, models.dir)

# store biomarkers in one file
save.df.to.file(biomarkers.synergy.res, file =
                paste0(biomarkers.dir, "biomarkers_per_synergy"))

# Biomarkers from section:
# `Common performance-related biomarkers`
biomarkers.res = add.performance.biomarkers(biomarkers.synergy.res, 
                                            biomarkers.mcc.active, 
                                            biomarkers.mcc.inhibited)

# prune nodes which are not found as biomarkers for any predicted synergy or
# for better model performance
biomarkers.res = biomarkers.res[, colSums(biomarkers.res != 0) > 0]

superheat(biomarkers.res, title.alignment = "center",
          title = paste0("Biomarker results (", cell.line, ")"),
          row.dendrogram = FALSE, col.dendrogram = TRUE, scale = FALSE,
          column.title = "Network Nodes", row.title = "Predicted synergies",
          heat.pal = c("tomato", "grey", "gold"),
          # grey   = not a biomarker
          # tomato = inhibited biomarker
          # gold   = active biomarker
          force.grid.vline = TRUE, force.grid.hline = TRUE,
          grid.vline.size = 0.1, grid.hline.size = 0.1, 
          legend.breaks = c(-1, 0, 1),
          left.label.col = "white", left.label.text.size = 4,
          bottom.label.col = "white",
          bottom.label.text.size = 2, bottom.label.text.angle = 90)
```

So, in general we observe that:

- A total of $43$ nodes were found as biomarkers (for at least one synergy)
- Most synergies have more than 10 biomarkers. Usually we wouldn't expect too 
many biomarkers that are directly related to the prediction of a specific synergy. 
The abudance of (false positive) biomarkers for some synergies (e.g. `PI-D1`) 
relates to the model classification method used, which does not incorporate in 
its internal logic that the prediction of other synergies than the ones used for 
the grouping itself can affect the biomarker results obtained from it
- We couldn't identify any biomarkers at the $0.7$ threshold for the `D1-P5` synergy
- All the performance-related biomarkers (PERF) were also observed as biomarkers
for the prediction of a specific synergy(ies) with the exception of the `RAF1` node
- There exist common biomarkers across different predicted synergies, e.g. `SGK3` 
is a common active biomarker across 3 synergistic drug combinations
- Most of the biomarkers found for this cell line have to be in the active state
- We observe that between the two predicted synergies `PI-D1` and `BI-PI` there
are conflicted biomarkers in the sense that they have to be in an more active 
state for the first synergy and more inhibited for the other (and vise versa)
- The synergies `JN-D1`, `PI-D1` and `PI-JN` share many common biomarkers as well 
as with the performance-related biomarkers

## References

1. Csardi G, Nepusz T (2006) The igraph software package for complex network research, InterJournal, Complex Systems 1695. [http://igraph.org](http://igraph.org)
2. Ognyanova, K. (2018) Network visualization with R. Retrieved from [www.kateto.net/network-visualization](http://www.kateto.net/network-visualization)
3. Barter Rebecca, Yu Bin (2017) Superheat: An R package for creating beautiful and extendable heatmaps for visualizing complex data. [https://arxiv.org/abs/1512.01524](https://arxiv.org/abs/1512.01524), [https://rlbarter.github.io/superheat/](https://rlbarter.github.io/superheat/)
