---
title: "Compare CLSS steady state vectors"
author: "[John Zobolas](https://github.com/bblodfon)"
date: "Last updated: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    css: style.css
    theme: united
    toc: false
    number_sections: false
    code_folding: hide
    code_download: true
---

```{r Render command, eval=FALSE, include=FALSE}
#rmarkdown::render(input = "compare_ss_clss.Rmd", output_format = "html_document", output_dir = "../../docs/cascade/cell-lines-2500/")
```

Loading libraries:
```{r Load libraries, message = FALSE}
library(emba)
library(usefun)
library(igraph)
library(dplyr)
library(DT)
```

Get the `.tsv` file with produced steady states for `AGS` using various computational methods:
```{r Input, comment=""}
# Get the compare steady state file 
ss.compare.file = paste0(getwd(), "/steady_state_comparison_AGS.tsv")

df = read.table(file = ss.compare.file, header = TRUE, fill = TRUE, sep = "\t")

# Get Cascade 2.0 Topology
topology.file = paste0(getwd(), "/topology")
net = construct_network(topology.file = topology.file, models.dir =  paste0(getwd(), "/AGS/models"))
cascade.nodes = V(net)$name

# check that only the genes (*_g) are not given a steady state value 
cascade.nodes[!cascade.nodes %in% df$Node]

# prune to the CASCADE 2.0 nodes only
df = df %>% filter(Node %in% cascade.nodes)

datatable(data = df, options = list(searching = FALSE, pageLength = 12))
```

**Compare AGS: PARADIGM only** new results and previous (See [README.md](https://github.com/bblodfon/gitsbe-model-analysis/blob/master/cascade/cell-lines-2500/steady-states/README.md) for the previous calculation method):
```{r, comment=""}
file = paste0(getwd(), "/steady-states/AGS_steady_state")
ss.df = read.table(file, sep = "\t", stringsAsFactors = FALSE)

paradigm.ss = ss.df[,2]
names(paradigm.ss) = ss.df[,1]
  
# change value to NA for nodes for which there was no activity found (dash)
paradigm.ss[paradigm.ss == "-"] = NA

paradigm.ss = prune_and_reorder_vector(paradigm.ss, as.character(df$Node))
paradigm.ss = sapply(paradigm.ss, as.integer)

df = cbind(df, paradigm.ss)

get_percentage_of_matches(df$paradigm, df$paradigm.ss)
```

**Compare AGS: PARADIGM+TF** new results and previous (See [README.md](https://github.com/bblodfon/gitsbe-model-analysis/blob/master/cascade/cell-lines-2500/steady-states/README.md) for the previous calculation method):
```{r, comment=""}
file = paste0(getwd(), "/AGS/steady_state")
ss.df = read.table(file, sep = "\t", stringsAsFactors = FALSE)

paradigm_tf.ss = ss.df[,2]
names(paradigm_tf.ss) = ss.df[,1]
  
# change value to NA for nodes for which there was no activity found (dash)
paradigm_tf.ss[paradigm_tf.ss == "-"] = NA

paradigm_tf.ss = prune_and_reorder_vector(paradigm_tf.ss, as.character(df$Node))
paradigm_tf.ss = sapply(paradigm_tf.ss, as.integer)

df = cbind(df, paradigm_tf.ss)

get_percentage_of_matches(df$paradigm_tf, df$paradigm_tf.ss)
```

**Compare AGS: PARADIGM only vs PARADIGM+TF** results (previous calculations, same as in the [fitness vs performance analysis](https://bblodfon.github.io/gitsbe-model-analysis/cascade/cell-lines-2500/performance_vs_fitness.html#input13)):
```{r, comment=""}
get_percentage_of_matches(df$paradigm.ss, df$paradigm_tf.ss)
```

**Compare AGS: PARADIGM only vs PARADIGM+TF** results (new calculation, provided by Miguel 20/11/2019):
```{r, comment=""}
get_percentage_of_matches(df$paradigm, df$paradigm_tf)
```

<div class="orange-box">
Results have changed for PARADIGM and PARADIGM+TF, but seem to have become more consistent!
</div>
</br>

**Who agrees more with literature-curated nodes for AGS?**

Filter to literature nodes:
```{r, comment=""}
df.filtered = df %>% filter(!is.na(literature))
```

Reporting results as in [fitness vs performance analysis](https://bblodfon.github.io/gitsbe-model-analysis/cascade/cell-lines-2500/performance_vs_fitness.html#input13)):
**Compare AGS: Literature (Gold) vs PARADIGM (previous) results**:
```{r, comment=""}
get_percentage_of_matches(df.filtered$literature, df.filtered$paradigm.ss)
```

**Compare AGS: Literature (Gold) vs PARADIGM+TF (previous) results**:
```{r, comment=""}
get_percentage_of_matches(df.filtered$literature, df.filtered$paradigm_tf.ss)
```

Comparing new results (based on Miguel's 20/11/2019 file):
**Compare AGS: Literature (Gold) vs PARADIGM (new) results**:
```{r, comment=""}
get_percentage_of_matches(df.filtered$literature, df.filtered$paradigm)
```

**Compare AGS: Literature (Gold) vs PARADIGM+TF (new) results**:
```{r, comment=""}
get_percentage_of_matches(df.filtered$literature, df.filtered$paradigm_tf)
```

<div class="orange-box">
Results seem to be consistenly random compared ot literature/Gold AGS steady state in all cases!
</div>
</br>