---
title: "Performance vs Fitness Model Analysis"
author: "[John Zobolas](https://github.com/bblodfon)"
date: "Last updated: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: united
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: false
    code_folding: hide
    code_download: true
---

```{r Render command, eval=FALSE, include=FALSE}
#rmarkdown::render(input = "./performance_vs_fitness.Rmd", output_format = "html_document", output_dir = "../../docs/cascade/cell-lines-2500/")
```

## Intro {-}

The purpose of this analysis is to find a correlation between the boolean models
**fitness to a steady state activity profile** and their **performance** in terms of the
number of *True Positive* (TP) synergies predicted and/or the overall *MCC score*
(Matthews Correlation Coefficient score). We want to show that **a closer fitness 
to the steady state suggests more predictive models**, corroborating thus our proof of concept of using an ensemble-based approach where models are trained towards a 
specific steady state signaling pattern for drug combination predictions.

The boolean model datasets we will use are in total $9$: one for each cell line 
of interest (8 cell lines) where the models were **fitted to a specific steady state** in each 
case and one for the so-called **random models** which were generated *randomly* in 
the sense that were fitted only to a proliferation state (simulations were done using 
the DrugLogics software modules `Gitsbe` and `Drabme`).

Each boolean model dataset constitues of:

- The **model predictions** file which has for each model the prediction for 
each drug combination tested (*0* = no synergy predicted, *1* = synergy 
predicted, *NA* = couldn't find stable states in either the drug combination 
inhibited model or in any of the two single-drug inhibited models)
- The **models stable state** (one per model). A **fitness score**
for each model can easily be calculated then by matching the model's stable 
state (which is something inherent in the boolean's model structure, a unique 
fixpoint attractor) with the steady state of interest, node per node.
A **higher fitness score** would mean a better match of a model's 
stable state to the cell line derived steady state (a perfect match would result 
in a fitness of 1).
- The **models link operators** which is a representation of the boolean equations
of each model. Each boolean equation is in the form: **Target *= (Activator OR Activator OR...) AND NOT (Inhibitor OR Inhibitor OR...)** and the difference between the models can be found in the 
*link operator* (*1* = 'OR NOT', *0* = 'AND NOT', or absent) which has been 
mutated (changed) through the genetic algorithm in `Gitsbe`. Note that the equations that do 
not have link operators are *the same for every model* and are thus discarded.
- The **observed synergies** file which lists the drug combinations that were 
observed as synergistic for each cell line.
- The **steady state** file which lists the network nodes (protein, gene, complexes
names, etc) and their activity value (0 or 1, representing an inhibited or active
node respectively). 
This input is provided per cell line and not for the random models since they are just trained to a profileration state.

```{r Load libraries, message = FALSE, echo = FALSE}
library(DT)
library(ggpubr)
library(emba)
library(usefun)
library(nnet)
library(pscl)
library(ComplexHeatmap)
library(circlize)
library(dplyr)
library(tibble)
library(Ckmeans.1d.dp)
library(RColorBrewer)
```

## Input {-}

First we load the cell-specific input data:
```{r Cell-specific Input}
# Cell Lines
cell.lines = c("A498", "AGS", "DU145", "colo205", "SW620", "SF295", "UACC62", "MDA-MB-468")

cell.line.dirs = sapply(cell.lines, function(cell.line) {
  paste0(getwd(), "/", cell.line)
})

# Model predictions
model.predictions.files = sapply(cell.line.dirs, function(cell.line.dir) {
  paste0(cell.line.dir, "/model_predictions")
})

model.predictions.per.cell.line = lapply(model.predictions.files, 
  function(file) {
    get_model_predictions(file)
  }
)

# Observed synergies
observed.synergies.files = sapply(cell.line.dirs, function(cell.line.dir) {
  paste0(cell.line.dir, "/observed_synergies")
})

observed.synergies.per.cell.line = lapply(observed.synergies.files, 
  function(file) {
    get_observed_synergies(file)
  }
)

# Models Stable State (1 per model)
models.stable.state.files = sapply(cell.line.dirs, function(cell.line.dir) {
  paste0(cell.line.dir, "/models_stable_state")
})

models.stable.state.per.cell.line = lapply(models.stable.state.files,
  function(file) {
    as.matrix(read.table(file, check.names = FALSE))
  }
)

# Models Link Operators
models.link.operator.files = sapply(cell.line.dirs, function(cell.line.dir) {
  paste0(cell.line.dir, "/models_link_operator")
})

models.link.operators.per.cell.line = lapply(models.link.operator.files,
  function(file) {
    as.matrix(read.table(file, check.names = FALSE))
  }
)

# the node names used in our analysis
node.names = colnames(models.stable.state.per.cell.line[[1]])

# Steady States
steady.state.files = sapply(cell.line.dirs, function(cell.line.dir) {
  paste0(cell.line.dir, "/steady_state")
})

steady.state.per.cell.line = lapply(steady.state.files,
  function(file) {
    ss.df = read.table(file, sep = "\t", stringsAsFactors = FALSE)
    steady.state = ss.df[,2]
    names(steady.state) = ss.df[,1]
    
    # change value to NA for nodes for which there was no activity found (dash)
    steady.state[steady.state == "-"] = NA
    
    # keep only the nodes that are included in the analysis
    steady.state = prune_and_reorder_vector(steady.state, node.names)
    
    # check
    stopifnot(names(steady.state) %in% node.names)
    
    # return an integer vector since the activity values are binarized (0,1)
    return(sapply(steady.state, as.integer))
  }
)
```

The random model input data:
```{r Random model Input}
random.dir = paste0(getwd(), "/random")
random.model.predictions = get_model_predictions(paste0(random.dir, "/model_predictions"))

random.models.stable.state = as.matrix(
  read.table(file = paste0(random.dir, "/models_stable_state"), check.names = FALSE)
)

random.models.link.operator =
  as.matrix(read.table(file = paste0(random.dir, "/models_link_operator"), check.names = FALSE))
```

## Model Analysis {-}

In order to find the number of true positive predicted synergies, MCC scores and fitness scores for each of the models in each of the 9 datasets, we use functions from the [emba](https://github.com/bblodfon/emba) R package.

### Cell-specific {-}

We find the MCC, TP and fitness values for each model per cell line (note
that each model's stable state in a specific cell line is matched against the 
steady state from that cell line):
```{r Cell-specific Models TP + MCC + fitness per cell line}
models.tp.per.cell.line = list()
models.mcc.per.cell.line = list()
models.fitness.per.cell.line = list()

for (cell.line in cell.lines) {
  model.predictions = model.predictions.per.cell.line[[cell.line]]
  observed.synergies = observed.synergies.per.cell.line[[cell.line]]
  number.of.drug.comb.tested = ncol(model.predictions.per.cell.line[[cell.line]])
  
  # Split model.predictions to positive (observed) and negative (non-observed) results
  observed.model.predictions =
    get_observed_model_predictions(model.predictions, observed.synergies)
  unobserved.model.predictions =
    get_unobserved_model_predictions(model.predictions, observed.synergies)
  
  # Count the predictions of the observed synergies per model (TP)
  models.tp.per.cell.line[[cell.line]] = calculate_models_synergies_tp(observed.model.predictions)
  
  # Calculate Matthews Correlation Coefficient (MCC) for every model
  models.mcc.per.cell.line[[cell.line]] = 
    calculate_models_mcc(observed.model.predictions,
                         unobserved.model.predictions,
                         number.of.drug.comb.tested)
  
  # Fitness per model contrasted to steady state from cell line
  steady.state = steady.state.per.cell.line[[cell.line]]
  models.stable.state = models.stable.state.per.cell.line[[cell.line]]
  models.fitness.per.cell.line[[cell.line]] = apply(
    models.stable.state[,names(steady.state)], 1, get_percentage_of_matches, steady.state
  )
}
```

### Random Models {-}

Finding the MCC, TP and fitness values for each model from the above result (note
that each model's stable state in a specific cell line is matched against the 
steady state from that cell line and that the random models' stable state data
does not change per cell line, i.e. same `random.models.stable.state` object):
```{r Random Models TP + MCC + fitness per cell line}
random.models.mcc.per.cell.line = list()
random.models.tp.per.cell.line = list()
random.models.fitness.per.cell.line = list()

for (cell.line in cell.lines) {
  observed.synergies = observed.synergies.per.cell.line[[cell.line]]
  number.of.drug.comb.tested = ncol(random.model.predictions)
  
  # Split model.predictions to positive (observed) and negative (non-observed) results
  observed.model.predictions =
    get_observed_model_predictions(random.model.predictions, observed.synergies)
  unobserved.model.predictions =
    get_unobserved_model_predictions(random.model.predictions, observed.synergies)
  
   # Count the predictions of the observed synergies per model (TP)
  random.models.tp.per.cell.line[[cell.line]] = 
    calculate_models_synergies_tp(observed.model.predictions)
  
  # Calculate Matthews Correlation Coefficient (MCC) for every model
  random.models.mcc.per.cell.line[[cell.line]] = 
    calculate_models_mcc(observed.model.predictions,
                         unobserved.model.predictions,
                         number.of.drug.comb.tested)
  
  # Fitness per model contrasted to steady state from cell line 
  steady.state = steady.state.per.cell.line[[cell.line]]
  random.models.fitness.per.cell.line[[cell.line]] = apply(
    random.models.stable.state[,names(steady.state)], 1, 
    get_percentage_of_matches, steady.state
  )
}
```

## Choosing best dataset for analysis {-}

We now want to **find the best dataset & cell line** for our subsequent analysis - that is to show the performance vs fitness correlation. 
The argument here is that we want to choose a boolean dataset that has a **large enough fitness value range** along with a large **TP and/or MCC value range**, since with smaller value ranges it would be harder to distinguish the difference of the estimated distributions of the fitness scores belonging to different performance classes (i.e. models' fitnesses that belong to different classification groups with the metric being either the number of TPs or the MCC score).

The next two tables will help us determine exactly which cell line and dataset
to use:
```{r Cell-specific Models TP + MCC + fitness per cell line stats}
cell.specific.model.data = matrix(data = NA, nrow = length(cell.lines), ncol = 11)
rownames(cell.specific.model.data) = cell.lines
colnames(cell.specific.model.data) = c("fitness range", "Min fitness", 
  "Max fitness", "Mean fitness", "Median fitness", "MCC range", "Min MCC", 
  "Max MCC", "Mean MCC", "Median MCC", "Max TP")

for (cell.line in cell.lines) {
  models.fitness = models.fitness.per.cell.line[[cell.line]]
  models.mcc = models.mcc.per.cell.line[[cell.line]]
  models.tp = models.tp.per.cell.line[[cell.line]]
  
  fit.summary = unclass(summary(models.fitness))
  mcc.summary = unclass(summary(models.mcc))
  max.tp = max(models.tp)
  
  cell.specific.model.data[cell.line, "fitness range"] = fit.summary["Max."] - fit.summary["Min."]
  cell.specific.model.data[cell.line, "Min fitness"] = fit.summary["Min."]
  cell.specific.model.data[cell.line, "Max fitness"] = fit.summary["Max."]
  cell.specific.model.data[cell.line, "Mean fitness"] = fit.summary["Mean"]
  cell.specific.model.data[cell.line, "Median fitness"] = fit.summary["Median"]
  
  cell.specific.model.data[cell.line, "MCC range"] = mcc.summary["Max."] - mcc.summary["Min."]
  cell.specific.model.data[cell.line, "Min MCC"] = mcc.summary["Min."]
  cell.specific.model.data[cell.line, "Max MCC"] = mcc.summary["Max."]
  cell.specific.model.data[cell.line, "Mean MCC"] = mcc.summary["Mean"]
  cell.specific.model.data[cell.line, "Median MCC"] = mcc.summary["Median"]
  
  cell.specific.model.data[cell.line, "Max TP"] = max.tp
}

# color columns
fit.breaks = quantile(cell.specific.model.data[,"fitness range"], probs = seq(.05, .95, .05), na.rm = TRUE)
fit.colors = round(seq(255, 40, length.out = length(fit.breaks) + 1), 0) %>%
  {paste0("rgb(255,", ., ",", ., ")")} # red
mcc.breaks = quantile(cell.specific.model.data[,"MCC range"], probs = seq(.05, .95, .05), na.rm = TRUE)
mcc.colors = round(seq(255, 40, length.out = length(mcc.breaks) + 1), 0) %>%
  {paste0("rgb(", ., ",255,", ., ")")} # green

caption.title = "Table 1: Fitness, MCC scores and TP (True positives) for the Cell-specific model predictions across 8 Cell Lines"
datatable(data = cell.specific.model.data, 
          options = list(dom = "t"), # just show the table
          caption = htmltools::tags$caption(caption.title, style="color:#dd4814; font-size: 18px")) %>% 
  formatRound(1:10, digits = 3) %>%
  formatStyle(columns = c("fitness range"), backgroundColor = styleInterval(fit.breaks, fit.colors)) %>%
  formatStyle(columns = c("MCC range"), backgroundColor = styleInterval(mcc.breaks, mcc.colors))
```

```{r Random Models TP + MCC + fitness per cell line stats}
random.model.data = matrix(data = NA, nrow = length(cell.lines), ncol = 11)
rownames(random.model.data) = cell.lines
colnames(random.model.data) = c("fitness range", "Min fitness", 
  "Max fitness", "Mean fitness", "Median fitness", "MCC range", "Min MCC", 
  "Max MCC", "Mean MCC", "Median MCC", "Max TP")

for (cell.line in cell.lines) {
  models.fitness = random.models.fitness.per.cell.line[[cell.line]]
  models.mcc = random.models.mcc.per.cell.line[[cell.line]]
  models.tp = random.models.tp.per.cell.line[[cell.line]]
  
  fit.summary = unclass(summary(models.fitness))
  mcc.summary = unclass(summary(models.mcc))
  max.tp = max(models.tp)
  
  random.model.data[cell.line, "fitness range"] = fit.summary["Max."] - fit.summary["Min."]
  random.model.data[cell.line, "Min fitness"] = fit.summary["Min."]
  random.model.data[cell.line, "Max fitness"] = fit.summary["Max."]
  random.model.data[cell.line, "Mean fitness"] = fit.summary["Mean"]
  random.model.data[cell.line, "Median fitness"] = fit.summary["Median"]
  
  random.model.data[cell.line, "MCC range"] = mcc.summary["Max."] - mcc.summary["Min."]
  random.model.data[cell.line, "Min MCC"] = mcc.summary["Min."]
  random.model.data[cell.line, "Max MCC"] = mcc.summary["Max."]
  random.model.data[cell.line, "Mean MCC"] = mcc.summary["Mean"]
  random.model.data[cell.line, "Median MCC"] = mcc.summary["Median"]
  
  random.model.data[cell.line, "Max TP"] = max.tp
}

# color columns
fit.breaks = quantile(random.model.data[,"fitness range"], probs = seq(.05, .95, .05), na.rm = TRUE)
fit.colors = round(seq(255, 40, length.out = length(fit.breaks) + 1), 0) %>%
  {paste0("rgb(255,", ., ",", ., ")")} # red
mcc.breaks = quantile(random.model.data[,"MCC range"], probs = seq(.05, .95, .05), na.rm = TRUE)
mcc.colors = round(seq(255, 40, length.out = length(mcc.breaks) + 1), 0) %>%
  {paste0("rgb(", ., ",255,", ., ")")} # green

caption.title = "Table 2: Fitness, MCC scores and TP (True positives) for the random model predictions across 8 Cell Lines"
datatable(data = random.model.data, 
          options = list(dom = "t"), # just show the table
          caption = htmltools::tags$caption(caption.title, style="color:#dd4814; font-size: 18px")) %>% 
  formatRound(1:10, digits = 3) %>%
  formatStyle(columns = c("fitness range"), backgroundColor = styleInterval(fit.breaks, fit.colors)) %>%
  formatStyle(columns = c("MCC range"), backgroundColor = styleInterval(mcc.breaks, mcc.colors))
```

The below box plots compare the MCC values and fitness scores across all cell lines between
the cell-specific models (trained to steady state) and the random ones (trained to proliferation):

```{r Combine data into one data frame}
num.of.models = nrow(random.models.stable.state)
data.list = list()

for (cell.line in cell.lines) {
  cell.line.vec = as.data.frame(rep(cell.line, num.of.models), stringsAsFactors = FALSE)
  models.mcc.cell.specific = remove_rownames(as.data.frame(models.mcc.per.cell.line[[cell.line]]))
  models.mcc.random        = remove_rownames(as.data.frame(random.models.mcc.per.cell.line[[cell.line]]))
  
  models.fitness.cell.specific = remove_rownames(as.data.frame(models.fitness.per.cell.line[[cell.line]]))
  models.fitness.random        = remove_rownames(as.data.frame(random.models.fitness.per.cell.line[[cell.line]]))
  
  data.list[[cell.line]] = bind_cols(cell.line.vec, models.mcc.cell.specific, 
    models.mcc.random, models.fitness.cell.specific, models.fitness.random)
}

data = bind_rows(data.list)
colnames(data) = c("cell.line", "MCC cell-specific", "MCC random", 
                   "fitness cell-specific", "fitness random")
```

```{r Boxplots: MCC and fitness values between cell-specific and random models, fig.width=9, warning=FALSE, cache=TRUE}
# Note the cell-specific models have NaN MCC values
ggboxplot(data, x = "cell.line", y = c("MCC cell-specific", "MCC random"),
          palette = brewer.pal(3, "Set1"), merge = "asis",
          xlab = "Cell Lines", ylab = "MCC values", add = "point", add.params = list(size = 0.5))
ggboxplot(data, x = "cell.line", y = c("fitness cell-specific", "fitness random"),
          palette = brewer.pal(3, "Set1"), merge = "asis",
          xlab = "Cell Lines", ylab = "Fitness values", add = "point", add.params = list(size = 0.5))
```

So, from the two tables and the two box plots above we conclude that:

- In general, the **random models offer a larger range of 
fitness values** when their stable states are matched against the steady state 
of each particular cell line. 
This is something we expected since these models weren't fitted to a 
specific cell-line steady state (meaning that they weren't chosen from the 
`Gitsbe` module as the 3 best from each simulation that match that steady state 
as best as possible) but rather to a more generic state of proliferation. 
Thus, they represent a set of models with larger variation in terms of 
structure (boolean model equations) compared to the cell-specific generated ones.
- **The cell-specific models have always a larger maximum fitness and median value** 
compared to the random ones for each respective cell line.
- In most of the cell lines, **the cell-specific models show an equal or less 
performance (smaller MCC values) than the random ones**

## R session info {-}

```{r session info, comment=""}
xfun::session_info()
```
